Exponentiation, anomalies and evanescence beyond one-loop

### 2023-05-12 AbelianTwoLoop.tex
added to Intro, and James' comments
updates

\abstract{
In this work we initiate a study of multi-loop amplitudes and quantum actions for effective gauge theories, including both the nonlinear sigma model (NLSM) and nonlinear supersymmetric gauge theories (DBIVA). We use a variety of on-shell methods that are well suited for studying two-loop order in these models. In computing these amplitudes, we identify a wide variety of rich physical structure. First, we find that the lead IR divergence of NLSM amplitudes exponentiates through two-loop order. Second, we compute the $U(1)$ anomaly in pure Born-Infeld theory, and find that the counterterms needed to cancel the anomaly can be constructed via a symmetric-structure double-copy. Furthermore, we show that the one-minus $(-+++)$ two-loop anomaly requires introducing an evanescent operator in a $D$-dimensional description of Born-Infeld theory. In addition to these pure photon counterterms, we show that the anomalous matrix elements that violate $U(1)$ duality invariance can be similarly cancelled by summing over internal $\mathcal{N}=4$ DBIVA superfields. In doing so, we find that $\mathcal{N}=4$ DBIVA amplitudes permits double-copy construction through two-loop order. This is shown by reproducing our unitarity based result with a loop-level double copy of $\mathcal{N}=4$ super-Yang-Mills numerators and our two-loop NLSM amplitudes. This result provides strong evidence for the color-dual nature of NLSM beyond one-loop. We conclude with an overview how $D$-dimensional four-photon counterterms can be constructed in generality with symmetric-structure double-copy, and outline a convenient way of counting evanescent operators at higher orders using Hilbert series as generating functions. }

\maketitle
\newpage
\section{Introduction}\label{sec:intro}
Here we initiate a study of effective field theory (EFT) at the multi-loop level using a combination of generalized unitarity \cite{UnitarityMethod, Fusing, BDKUniarityReview} and the double-copy construction \cite{BCJ, BCJLoop}. The EFTs that we choose to study in this work are the nonlinear sigma model (NLSM) and nonlinear supersymmetric gauge theories, known formally as Dirac-Born-Infeld-Volkov-Akulov (DBIVA) theories. The motivation for this work is two-fold. First, perturbative calculations in this class of effective field theories are significantly simpler than the typical multi-loop calculations of renormalizable theories, like quantum-chromodynamics (QCD) and super-Yang-Mills (sYM). As we'll show, the multi-loop amplitudes for these theories allow us to recycle many of the $D$-dimensional integration tool that are so powerful at one-loop order. This allows us to compute a large catalog of $D$-dimensional two-loop amplitudes, which provides fertile ground for cultivating insights about the multi-loop structure of effective field theories. The perturbative depth of our calculations sheds light on (1) the exponential structure of IR divergences for NLSM amplitudes through two-loops, (2) the anomalous behavior $U(1)$ duality invariance through two-loop order in DBIVA theories, (3) and the emergence of evanescent operators relevant for anomaly cancellation in pure Born-Infeld theory. We believe these surprisingly rich physical structures could serve as a theoretical laboratory for future studies of the interplay between effective field theory operators and perturbative calculations in quantum field theory.

Second, despite the success of applying color-kinematics duality to $\mathcal{N}=4$ super-Yang-Mills, for which color-dual integrands are known through four-point four-loop \cite{Neq44np, GravityFour} and seven-point one-loop \cite{Edison:2020uzf,Edison:2022jln}, beyond these representations very little is known about the multi-loop level for other models in the web of theories. For a comprehensive review of color-dual representations, see refs. \cite{BCJreview, Bern:2022wqg, Adamo:2022dcm}. Presently, the state-of-the-art for nonlinear sigma model (NLSM) color-dual integrands come from the Y-Z model of Cheung and Shen \cite{Cheung2016prv}, while $D$-dimensional integrands for pure Yang-Mills have only been identified through five-point one-loop \cite{He:2017spx}. At present, there are no known $D$-dimensional representations for either of these theories at two-loop that globally manifest the duality between color and kinematics, despite attempts in the literature \cite{OneTwoLoopPureYMBCJ, Geyer:2019hnn}. Furthermore, similar bottlenecks exists for less-than-maximal $\mathcal{N}\leq 2$ sYM \cite{Mogull:2015adi}. While there have been a number of recent developments in constructing manifestly color-dual Feynman rules \cite{Chen:2019ywi,Chen:2021chy,Brandhuber:2021bsf,Cheung:2021zvb,Ben-Shahar:2021zww,Cheung:2022mix, Ben-Shahar:2022ixa}, a precise definition the kinematic algebra off-shell remains elusive. Considering the known tree-level double-copy construction of NLSM and DBIVA \cite{Cachazo:2014xea}, all of the amplitudes we compute should in principle participate in a double-copy formulation of our results. At the very least, the amplitudes we compute will serve as important checks on future breakthroughs in multi-loop studies of color-dual representations. In addition to traditional double-copy construction, we find that the quantum effective actions generated by these loop effect of this paper permit a rather elegant construction in terms of \textit{symmetric-structure double-copy}, introduced in a recent work by the authors \cite{Carrasco:2022jxn}.

To begin, we provide a brief review of the types of models studied in this work, namely NLSM and DBIVA. We call such theories \textit{abelianized} because the appear naturally in in the field theory limit of the \textit{abelianized open superstring} \cite{Green:1982sw}. Tree-level amplitudes for the open superstring (OSS) can be constructed from a double copy \cite{Broedel:2013tta} between $\mathcal{N}=N_f$ super-Yang-Mills (sYM) and $Z$-theory amplitudes \cite{Mafra:2011nv,Mafra:2011nw}, as follows,
\begin{equation}
A^{\text{OSS}}= Z \otimes A^{ \text{sYM}}
\end{equation}
where $\otimes$ implements the field theory double copy \cite{KLT, BCJ}. To recover the abelian sector, one simply sums over all color orderings on the side of the bi-colored $Z$-theory amplitudes that obey string monodromy relations. In the field theory limit, the observables for so-called \textit{abelianized} $Z$-theory are simply the amplitudes generated by the NLSM Lagrangian \cite{Carrasco:2016ldy}. That is, given a $n$-point amplitude in $Z$-theory, with field theory ordering $a=(a_1,a_2,...,a_n)$ and string theoretic color ordering $A=(A_1,A_2,...,A_n)$, one finds the following relation:
\begin{equation}
A^{\text{NLSM}}{(a_1,a_2,...,a_n)} \equiv \lim_{\alpha' \to 0} (\alpha')^{2-n}  Z_{\times}{(a_1,a_2,...,a_n)}
\end{equation}
where $Z_{\times}$ are abelianized $Z$-theory amplitudes of \cite{}, defined by summing all possible orderings of Chan-Paton factors:
\begin{equation}
Z_{\times}(a_1,a_2,...,a_n)\equiv \sum_{A \in S^{n-1}} Z_{(A_1,...,A_{n-1},n)}{(a_1,a_2,...,a_n)}
\end{equation}
Since the field theory limit of abelian $Z$-theory produces NLSM amplitudes, the field theory limit of the abelianized OSS, gives rise to DBIVA observables at leading order:
\begin{equation}
\boxed{\begin{split} \lim_{\alpha' \to 0 }Z^{\text{OSS}}_{\times}&= A^{\text{NLSM}} \otimes A^{\text{BAS}}\equiv A^{\text{NLSM}}
\\
\Rightarrow \quad\lim_{\alpha' \to 0 }A^{\text{OSS}}_{\times}&= A^{\text{NLSM}} \otimes A^{ \text{sYM}}\equiv A^{\text{DBIVA}}
\end{split}}
\end{equation}
These two classes of amplitudes, NLSM and DBIVA, are the observables that we will consider in the multi-loop studies of this work. While not important for our results, details on the bi-colored $Z$-theory amplitudes these string-theoretic double-copy constructions can be found in \cite{Mafra:2011nv,Mafra:2011nw,Broedel:2013tta} and references therein.

By far the most important feature of these amplitudes relevant for our analysis is that they are generated by purely \textit{even-point} contacts. That is, the effective actions of abelianized theories are consistent with the following interaction Lagrangian:
\begin{equation}
\mathcal{L}^{\text{abelian}}_{\text{int.}} = \sum_{k=2} \alpha'^{n-2} \mathcal{O}^{(2n)}_{2k} 
\end{equation}
where $2k$ counts the field content of the effective operator, and $2n$ counts the mass-dimension determined by the number of fields and derivatives contained in the operator. The dimensionful scale, $\alpha'$, carries mass dimension $[\alpha']=-2$ in $d=4$.

The even-point nature of this class of theories make them comparatively easy to compute relative to their odd-multiplicity cousins. Indeed, the observation that will drive many of the multi-loop results is the following: \textit{amplitudes for abelianized theories at next-to-next-to-leading-order (NNLO) in $\alpha'$ are recursively one-loop}. To gain some familiarity with why this is the case, consider the interactions needed to construct two loop amplitudes for a simple $\varphi^{2k}$-scalar theory:
\begin{equation}\label{eq:evenPointL}
\mathcal{L}_{\varphi^{2k}} = \frac{1}{2}(\partial \varphi)^2 + c_4 \varphi^4 + c_6 \varphi^6+ c_8 \varphi^8+\cdots 
\end{equation}
While the Feynman rules for this theory are trivial, unlike those needed for the vectors and fermions in the DBIVA supermultiplet and the derivatively coupled pions in NLSM, the Feynman diagrams needed to compute amplitudes are identical to those of $A^{\text{OSS}}_{\times}$. Indeed, the two-loop amplitude for $\varphi^{2k}$ theory is generated simply by evaluating the following set of scalar 1PI integrals:

@@

In addition to the perturbative structure of these theories, DBIVA amplitudes have an additional special property that we will study in his work. At tree-level, these amplitudes exhibit $U(1)$ \textit{duality invariance} \cite{Bossard:2012xs,Carrasco:2013ypa,Cachazo:2014xea,Bern:2017rjw,Novotny:2018iph,Elvang:2020kuj}, such that the effect of rotating field strengths, $F^{\mu\nu}$, into the dual-fields, $G^{\mu\nu}$, defined below,
\be
G^{\mu\nu} = \epsilon^{\mu\nu\rho\sigma}\frac{\partial \mathcal{L}}{\partial F^{\mu\nu}}
\ee
leaves the equations of motion invariant \cite{Gibbons:1995ap,Babaei-Aghbolagh:2013hia}. This dynamical symmetry gives rise to a 4D helicity selection rule \cite{Novotny:2018iph} whereby tree-level matrix elements outside the split-helicity sector vanish on-shell
\be
\mathcal{M}^{\text{DBIVA}}_{(n_{-}, n_{+})} =0 \qquad \Leftrightarrow \qquad n_{-}\neq n_{+}
\ee
where $n_{(+)}$ and $n_{(-)}$ is the number of external positive and negative helicity photons, respectively. For $\mathcal{N}=0$ pure Born-Infeld theory \cite{Born:1934gh,Schrodinger:1935oqa}, this symmetry is violated at four-point one-loop \cite{Elvang:2019twd} in the all-plus $(++++)$ helicity sector \cite{Elvang:2020kuj}. In the text, we will show how his anomaly is propagated to two-loop, and demonstrate that $U(1)$ duality invariance is further broken in the $(-+++)$ configuration beyond one-loop. However, in the case of supersymmetric DBIVA, the $U(1)$ symmetry is promoted an $R$-symmetry, and is protected perturbatively by supersymmetric Ward Identities \cite{Heydeman:2017yww}. We will show this explicitly in the case of $\mathcal{N}=4$ DBIVA through two-loop.

The outline of the paper is as follows: in \sect{sec:IntMethods}, we will provide an overview of the on-shell methods and integration methods needed to probe the multi-loop physics studied in this paper. In \sect{sec:Loops}, we present our results, beginning with a warm-up calculation \sect{sec:NLSMU} where we construct the integrands for NLSM through NNLO in the effective coupling, and compute the fully integrated amplitudes. We show that in $D=2-2\epsilon$, the leading IR divergences of the theory exponentiate. With the scaffolding of $D$-dimensional integration in hand, we then compute two-loop amplitudes for $\mathcal{N}=4$ DBIVA theory and pure-photon Born-Infeld theory in \sect{sec:DBIU}. There we compute the anomalies present beyond one-loop order.  Then, using the NLSM integrands of \sect{sec:NLSMU}, we perform a multi-loop double-copy to $\mathcal{N}=4$ DBIVA observables with the color-dual basis numerators available in the literature in \sect{sec:DBIvDC}. After computing the catalog of two-loop amplitudes, we study the construction of quantum effective actions in \sect{sec:Actions} as they relate to the anomalous matrix elements present in two-loop Born-Infeld amplitudes. We demonstrate in \sect{sec:Anomalies} Actions that the one-minus anomaly at two-loop requires the introduction of an evanescent operator at $\mathcal{O}(\alpha'^4)$. Then in \sect{sec:ActionDC} we take the opportunity to discuss the interplay between symmetric-structure double copy, and higher-spin modes studied in a recent work by the authors. Finally, we lay out a Hilbert series framework for counting evanescent operators at general orders in mass-dimension. To conclude, in \sect{sec:EOpsCounting} we discuss many directions of future study and comment on the insights gained from this study. 

@@

To perform the loop-level calculations in this paper, we will construct the integrand in $D$-dimensions using generalized unitarity cuts \cite{UnitarityMethod}. In the kinematic regime where the internal loop momenta go on-shell, the residue is a product of on-shell tree level diagrams, summed over all physical states that can cross the cut. For example, the kinematic numerator of the one-loop integrand $\mathcal{I}_{\text{1-loop}}$ of pure Born-Infeld theory should satisfy the following constraint:

@@

It is well known that one-loop amplitudes are spanned by a small basis of scalar integrals \cite{Forde:2007mi,Badger:2008cm,ElvangHuangReview}. This property can be exposed via a $D$-dimensional integral reduction algorithm due to Passarino and Veltman \cite{Passarino:1978jh}. However, as we will demonstrate below, the Passarino-Veltman reduction algorithm can also be applied to the \textit{two-loop} integrals needed for our calculation since they are all \textit{recursively one-loop}. We will now demonstrate this property of recursively one-loop integrals in the next section.

@@

\begin{align}
\scaleIntAvectorODD{}{}{}{}{$\mathcal{O}^{\text{ev.}}$}{$\mathcal{O}^{\text{ex.}}$} &= \frac{(D_s-4)(D_s-3)}{8(D_s-1)}s_{12}^4 I_2(k_{12}) f_{12}f_{34}
\\
& = \frac{i s_{12}^4}{192\pi^2} f_{12}f_{34} +\mathcal{O}(\epsilon)
\end{align}

where $I_2(k_{12})$ is the scalar bubble integral in the $s_{12}$-channel, and in the second line we have plugged in $D=4-2\epsilon$. Thus far, this is exactly what we should expect. Since the operator $\mathcal{O}^{\text{ev.}}$ vanishes in $D=4$, the 4-dimensional cut above must vanish. The Optical Theorem thus disallows imaginary parts from logarithms that would appear post integration. This physical constraint is imposed by the factor of $(D-4)$, which absorbs the divergence of $I_2(k_{12})$ and pushes the logarithm to be subleading at $\mathcal{O}(\epsilon)$.

@@

where $g_i \equiv 2 k_{i-1}^\mu F^{\mu\nu}_i k^{\nu}_{i+1}$. While not obvious when expressed in this form, the numerator of the above tensor structure is proportional to the permutation invariant in the denominator, $\mathcal{O}^{F^2F^2}_{(0,2)} -g_1g_2g_3g_4\sim \sigma_3$, and thus it is local by construction. This set of operators completely spans the space of $D$-dimensional permutation invariant photon effective operators, thereby allowing us to write an on-shell effective amplitude 

@@

\item \textbf{Projection}: Project $D$-dimensional tensor structures along defined 4D spinor helicity states. This will elucidate where anomalous matrix elements appear for non supersymmetric DBIVA, which has a classical conserved $U(1)$ symmetry. 

@@

Before diving into the fully dressed fermion and vector integrals of DBIVA theories, we will begin with an instructive example of NLSM pions through two-loop order. These amplitudes will serve as the scaffolding constructing $\mathcal{N}=4$ DBIVA integrands using the double copy. As we will demonstrate, an exciting realization of our results is that the loop-level double copy construction of $\mathcal{N}=4$ DBIVA is consistent with the results that we compute via unitarity. This is the first verification of the double-copy construction for multi-loop amplitudes for non-gravitational theories. Furthermore, it suggests that there should exist a color-dual representation of two-loop NLSM integrands, which thus far has evaded identification in the literature. We see finding such a color-dual pion integrand as an exciting direction of future study.   

@@

\mathcal{A}^{\text{NLSM}}_n = \sum_{\sigma \in S^{n-2}} C^{\text{ladder}}_{(1\sigma n)} A^{\text{NLSM}}_{(1\sigma n)}

@@

With this, we are finally prepared to start calculating! As we'll see, the $D$-dimensional calculations for pions are quite simple -- the integrand expressions are only a few lines of dot products. Guided by the simplicity of NLSM, we will label each step in the process according to the outline provided at the beginning of the section. While the complexity of the calculation will necessarily increase for DBIVA theories, the general procedure will remain the same.

>>>>

\abstract{
In this work we initiate a study of multi-loop amplitudes and quantum actions for effective field theories, including both the nonlinear sigma model (NLSM) and nonlinear supersymmetric gauge theories (DBIVA). We use a variety of on-shell methods that are well suited for studying two-loop order in these models. In computing these amplitudes, we identify a wide variety of rich physical structure. First, we find that the lead IR divergence of NLSM amplitudes exponentiates through two-loop order. Second, we compute the $U(1)$ anomaly in pure Born-Infeld theory, and find that the counterterms needed to cancel the anomaly can be constructed via a symmetric-structure double-copy. Furthermore, we show that the one-minus $(-+++)$ two-loop anomaly requires introducing an evanescent operator in a $D$-dimensional description of Born-Infeld theory. In addition to these pure photon counterterms, we show through explicit calculation that the anomalous matrix elements that violate $U(1)$ duality invariance can be similarly cancelled by summing over internal $\mathcal{N}=4$ DBIVA superfields. In doing so, we find that $\mathcal{N}=4$ DBIVA amplitudes permit double-copy construction through two-loop order. This is shown by reproducing our unitarity based result with a loop-level double copy of $\mathcal{N}=4$ super-Yang-Mills numerators and our two-loop NLSM amplitudes. This result provides strong evidence for the color-dual nature of NLSM beyond one-loop. We conclude with an overview of how $D$-dimensional four-photon counterterms can be constructed in generality with the symmetric-structure double-copy, and outline a convenient way of counting evanescent operators at higher orders using Hilbert series as generating functions. }

\maketitle
\newpage
\section{Introduction}\label{sec:intro}
Here we initiate a study of effective field theory (EFT) at the multi-loop level using a combination of generalized unitarity \cite{UnitarityMethod, Fusing, BDKUniarityReview} and the double-copy construction \cite{BCJ, BCJLoop}. The EFTs that we choose to study in this work are the nonlinear sigma model (NLSM) and nonlinear supersymmetric gauge theories, known formally as Dirac-Born-Infeld-Volkov-Akulov (DBIVA) theories. The motivation for this work is three-fold. First, perturbative calculations in NLSM and DBIVA effective field theories are significantly simpler than the typical multi-loop calculations of renormalizable theories, like quantum-chromodynamics (QCD) and super-Yang-Mills (sYM). As we will show, the multi-loop amplitudes for these theories allow us to recycle many of the $D$-dimensional integration tool that are so powerful at one-loop order. This allows us to compute a large catalog of $D$-dimensional two-loop amplitudes, which provides fertile ground for cultivating insights about the multi-loop structure of effective field theories. The perturbative depth of our calculations sheds light on (1) the exponential structure of IR divergences for NLSM amplitudes through two-loops, (2) the anomalous behavior $U(1)$ duality invariance through two-loop order in DBIVA theories, (3) and the emergence of evanescent operators relevant for anomaly cancellation in pure Born-Infeld theory. We believe these surprisingly rich physical structures could serve as a theoretical laboratory for future studies of the interplay between effective field theory operators and perturbative calculations in quantum field theory.

The second motivation is that very little is known about color-kinematics duality at multi-loop level for generic models in the web of theories \cite{BCJreview}. For a comprehensive review of color-dual representations, see refs. \cite{BCJreview, Bern:2022wqg, Adamo:2022dcm} and references therein. While there has been tremendous success of applying color-kinematics duality to $\mathcal{N}=4$ super-Yang-Mills, for which color-dual integrands are known through four-point four-loop \cite{Bern:2012uf, Neq44np, GravityFour} and seven-point one-loop \cite{Bjerrum-Bohr:2013iza,Edison:2020uzf,Edison:2022jln}, there are many obstructions for generic gauge/gravity theories. Presently, the state-of-the-art for nonlinear sigma model (NLSM) color-dual numerators comes from the $Y$-$Z$ model of Cheung and Shen \cite{Cheung2016prv}, while $D$-dimensional integrands for pure Yang-Mills have only been identified through five-point one-loop \cite{He:2017spx}. At present, there are no known $D$-dimensional representations for either of these theories at two-loop that globally manifest the duality between color and kinematics, despite attempts in the literature \cite{OneTwoLoopPureYMBCJ, Mogull:2015adi, Geyer:2019hnn}. Furthermore, similar bottlenecks exists for less-than-maximal $\mathcal{N}\leq 2$ sYM \cite{Johansson:2017bfl}. While there have been a number of recent developments in constructing manifestly color-dual Feynman rules \cite{Chen:2019ywi,Chen:2021chy,Brandhuber:2021bsf,Cheung:2021zvb,Ben-Shahar:2021zww,Cheung:2022mix, Ben-Shahar:2022ixa}, a precise definition the kinematic algebra off-shell remains elusive. Considering the known tree-level double-copy construction of NLSM and DBIVA \cite{Cachazo:2014xea}, all of the amplitudes we compute should in principle participate in a double-copy formulation of our results. At the very least, the amplitudes we compute will serve as important checks on future breakthroughs in multi-loop studies of color-dual representations. In addition to traditional double-copy construction, we find that the quantum effective actions generated by these loop effect of this paper permit a rather elegant construction in terms of \textit{symmetric-structure double-copy}, introduced in a recent work by the authors \cite{Carrasco:2022jxn}.

Lastly, DBIVA effective field theories touch a wide range research areas at the forefront of high energy physics. Of more formal interest, is the presence of quantum anomalies that violate the $U(1)$ duality invariance, which the theory enjoys at tree-level \cite{Elvang:2020kuj}. The presence of these $U(1)$ anomalies in gravity is closely linked to ultraviolet divergences computed at loop-level \cite{Bern:2007xj,Carrasco:2013ypa,Craig:2019zkf,Monteiro:2022nqt}. It has been argued that cancelling these anomalous matrix elements with $R^n$ counterterms can lead to enhanced UV cancellations \cite{Bern:2017tuc,Bern:2017rjw,Bern:2019isl}. This relationship has been demonstrated both for pure Einstein-Hilbert gravity \cite{Goroff:1985sz}, and also for less than maximal $\mathcal{N}\leq 4$ supergravity \cite{Bern:2013uka}. While DBIVA theories are themselves are ultraviolet divergent in $D=4$, due to their simplicity they serve as an essential laboratory for probing anomaly cancellation at high loop order. Moreover, in addition to their formal theory relevance, DBI has garnered wide phenomonological interest in cosmology, both for sourcing non-gaussianities in CMB bispectrum \cite{Alishahiha:2004eh,Creminelli:2005hu,Fergusson:2008ra}, and for their consistency with the observed CMB tensor mode suppression \cite{Carrasco:2015pla,Carrasco:2015rva,Carrasco:2015uma,BICEP:2021xfz,Kallosh:2021mnu}. Thus, with inflationary data on the horizon, understanding the perturbative structure of DBI could be particularly relevant for modeling early universe quantum fluctuations.

To begin, we provide a brief review of the types of models studied in this work, namely NLSM and DBIVA. While NLSM in general carries nonabelian color structure, we call such theories \textit{abelianized} because the both appear naturally in in the field theory limit of the \textit{abelianized open superstring} \cite{Green:1982sw}. Tree-level amplitudes for the open superstring (OSS) can be constructed from a double copy \cite{Broedel:2013tta} between $\mathcal{N}=N_f$ super-Yang-Mills (sYM) and $Z$-theory amplitudes \cite{Mafra:2011nv,Mafra:2011nw}, as follows,
\begin{equation}
A^{\text{OSS}}= Z \otimes A^{ \text{sYM}}
\end{equation}
where $\otimes$ implements the field theory double copy \cite{KLT, BCJ}. To recover the abelian sector, one simply sums over all color orderings on the side of the bi-colored $Z$-theory amplitudes that obey string monodromy relations. In the field theory limit, the observables for so-called \textit{abelianized} $Z$-theory are simply the amplitudes generated by the NLSM Lagrangian \cite{Carrasco:2016ldy}. That is, given a $n$-point amplitude in $Z$-theory, with field theory ordering $a=(a_1,a_2,...,a_n)$ and string theoretic color ordering $A=(A_1,A_2,...,A_n)$, one finds the following relation:
\begin{equation}
A^{\text{NLSM}}{(a_1,a_2,...,a_n)} \equiv \lim_{\alpha' \to 0} (\alpha')^{2-n}  Z_{\times}{(a_1,a_2,...,a_n)}
\end{equation}
where $Z_{\times}$ are abelianized $Z$-theory amplitudes of \cite{Carrasco:2016ldy}, defined by summing all possible orderings of Chan-Paton factors:
\begin{equation}
Z_{\times}(a_1,a_2,...,a_n)\equiv \sum_{A \in S^{n-1}} Z_{(A_1,...,A_{n-1},n)}{(a_1,a_2,...,a_n)}
\end{equation}
Since the field theory limit of abelian $Z$-theory produces NLSM amplitudes, the field theory limit of the abelianized OSS, gives rise to DBIVA observables at leading order:
\begin{equation}
\boxed{\begin{split} \lim_{\alpha' \to 0 }Z^{\text{OSS}}_{\times}&= A^{\text{NLSM}} \otimes A^{\text{BAS}}\equiv A^{\text{NLSM}}
\\
\Rightarrow \quad\lim_{\alpha' \to 0 }A^{\text{OSS}}_{\times}&= A^{\text{NLSM}} \otimes A^{ \text{sYM}}\equiv A^{\text{DBIVA}}
\end{split}}
\end{equation}
These two classes of amplitudes, NLSM and DBIVA, are the observables that we will consider in the multi-loop studies of this work. While not important for our results, details on the bi-colored $Z$-theory amplitudes these string-theoretic double-copy constructions can be found in \cite{Mafra:2011nv,Mafra:2011nw,Broedel:2013tta} and references therein.

By far the most important feature of these amplitudes relevant for our analysis is that they are generated by purely \textit{even-point} contacts. That is, the effective actions of abelianized theories are consistent with the following interaction Lagrangian:
\begin{equation}
\mathcal{L}^{\text{abelian}}_{\text{int.}} = \sum_{k=2} \alpha'^{n-2} \mathcal{O}^{(2n)}_{2k} 
\end{equation}
where $2k$ counts the field content of the effective operator, and $2n$ counts the mass-dimension determined by the number of fields and derivatives contained in the operator. The dimensionful scale, $\alpha'$, carries mass dimension $[\alpha']=-2$ in $D=4$.

The even-point nature of this class of theories make them comparatively easy to compute relative to their arbitrary-multiplicity cousins. Indeed, the observation that will drive many of the multi-loop results is the following: \textit{amplitudes for abelianized theories at next-to-next-to-leading-order (NNLO) in $\alpha'$ are recursively one-loop}. To gain some familiarity with why this is the case, consider the interactions needed to construct two loop amplitudes for a simple $\varphi^{2k}$-scalar theory:
\begin{equation}\label{eq:evenPointL}
\mathcal{L}_{\varphi^{2k}} = \frac{1}{2}(\partial \varphi)^2 + c_4 \varphi^4 + c_6 \varphi^6+ c_8 \varphi^8+\cdots 
\end{equation}
While the Feynman rules for this theory are trivial, unlike those needed for the vectors and fermions in the DBIVA supermultiplet and the derivatively coupled pions in NLSM, the Feynman diagrams needed to compute amplitudes are identical to those of $A^{\text{OSS}}_{\times}$. Indeed, the two-loop amplitude for $\varphi^{2k}$ theory is generated simply by evaluating the following set of scalar 1PI integrals:

@@

In addition to the perturbative structure of these theories, DBIVA amplitudes have an additional special property that we will study in his work. At tree-level, these amplitudes exhibit $U(1)$ \textit{duality invariance} \cite{Bossard:2012xs,Carrasco:2013ypa,Cachazo:2014xea,Bern:2017rjw,Novotny:2018iph,Elvang:2020kuj}, such that the effect of rotating field strengths, $F^{\mu\nu}$, into the dual-fields, $G^{\mu\nu}$, defined below,
\be
G^{\mu\nu} = \epsilon^{\mu\nu\rho\sigma}\frac{\partial \mathcal{L}}{\partial F^{\mu\nu}}
\ee
leaves the equations of motion invariant \cite{Gibbons:1995ap,Babaei-Aghbolagh:2013hia}. This dynamical symmetry gives rise to a 4D helicity selection rule \cite{Novotny:2018iph} whereby tree-level matrix elements outside the split-helicity sector vanish on-shell
\be
\mathcal{M}^{\text{DBIVA}}_{(n_{-}, n_{+})} =0 \qquad \Leftrightarrow \qquad n_{-}\neq n_{+}
\ee
where $n_{(+)}$ and $n_{(-)}$ is the number of external positive and negative helicity photons, respectively. For $\mathcal{N}=0$ pure Born-Infeld theory \cite{Born:1934gh,Schrodinger:1935oqa}, this symmetry is violated at four-point one-loop \cite{Elvang:2019twd} in the all-plus $(++++)$ helicity sector \cite{Elvang:2020kuj}. In the text, we will show how his anomaly is propagated to two-loop, and demonstrate that $U(1)$ duality invariance is further broken in the $(-+++)$ configuration beyond one-loop. However, in the case of supersymmetric DBIVA, the $U(1)$ symmetry is promoted to an $R$-symmetry, and is protected perturbatively to all orders by supersymmetric Ward Identities \cite{Heydeman:2017yww}. We will show this explicitly in the case of $\mathcal{N}=4$ DBIVA through two-loop.

The outline of the paper is as follows: in \sect{sec:IntMethods}, we will provide an overview of the on-shell methods and integration methods needed to probe the multi-loop physics studied in this paper. In \sect{sec:Loops}, we present our results, beginning with a warm-up calculation in \sect{sec:NLSMU} where we construct the integrands for NLSM through NNLO in the effective coupling, and compute the fully integrated amplitudes. We show that in $D=2-2\epsilon$, the leading IR divergences of the theory exponentiate. With the scaffolding of $D$-dimensional integration in hand, we then compute two-loop amplitudes for $\mathcal{N}=4$ DBIVA theory and pure-photon Born-Infeld theory in \sect{sec:DBIU}. There we compute the anomalies present beyond one-loop order.  Then, using the NLSM integrands of \sect{sec:NLSMU}, we perform a multi-loop double-copy to $\mathcal{N}=4$ DBIVA observables with the color-dual basis numerators available in the literature in \sect{sec:DBIvDC}. After computing the catalog of two-loop amplitudes, we study the construction of quantum effective actions in \sect{sec:Actions} as they relate to the anomalous matrix elements present in two-loop Born-Infeld amplitudes. We demonstrate in \sect{sec:Anomalies} that the one-minus anomaly at two-loop requires the introduction of an evanescent operator at $\mathcal{O}(\alpha'^4)$. Then in \sect{sec:ActionDC} we take the opportunity to discuss the interplay between symmetric-structure double copy, and higher-spin modes studied in a recent work by the authors. Finally, we lay out a Hilbert series framework for counting evanescent operators at general orders in mass-dimension. To conclude, in \sect{sec:EOpsCounting} we discuss many directions of future study and comment on the insights gained from this study. 

@@

To perform the loop-level calculations in this paper, we will construct the integrand in $D$-dimensions using generalized unitarity cuts \cite{Bern:1995db,Bern:1996ja}. In the kinematic regime where the internal loop momenta go on-shell, the residue is a product of on-shell tree level diagrams, summed over all physical states that can cross the cut. For example, the kinematic numerator of the one-loop integrand $\mathcal{I}_{\text{1-loop}}$ of pure Born-Infeld theory should satisfy the following constraint:

@@

It is well known that one-loop amplitudes are spanned by a small basis of scalar integrals \cite{Forde:2007mi,Badger:2008cm,ElvangHuangReview}. This property can be exposed via a $D$-dimensional integral reduction algorithm due to Passarino and Veltman \cite{Passarino:1978jh}. However, as we will demonstrate below, the Passarino-Veltman reduction algorithm can also be applied to the \textit{two-loop} integrals needed for our calculation since they are all \textit{recursively one-loop}. While there are a number of tensor reduction algorithms available in the literature \cite{Anastasiou:2004vj,vonManteuffel:2012np,Smirnov:2014hma,vonManteuffel:2014ixa,Smirnov:2019qkx,Smirnov:2020quc,Usovitsch:2020jrk,Maierhofer:2018gpa}, our ability to use Passarino-Veltman at two-loop will prove useful for terms like $(\varepsilon_i \cdot l_j)$, which include formal polarizations. We will articulate this in more detail in \sect{sec:2loopN4U}. Now we will demonstrate this property of recursively one-loop integrals in the next section.

@@

\begin{align}
\scaleIntAvectorODD{}{}{}{}{$\mathcal{O}^{\text{ev.}}$}{$\mathcal{O}^{\text{ex.}}$} &= \frac{(D_s-4)(D_s-3)}{8(D_s-1)}s_{12}^4 I_2(k_{12}) f_{12}f_{34}
\\
& = \frac{i s_{12}^4}{192\pi^2} f_{12}f_{34} +\mathcal{O}(\epsilon)
\end{align}
where $I_2(k_{12})$ is the scalar bubble integral in the $s_{12}$-channel, and in the second line we have plugged in $D=4-2\epsilon$. Thus far, this is exactly what we should expect. Since the operator $\mathcal{O}^{\text{ev.}}$ vanishes in $D=4$, the 4-dimensional cut above must vanish. The Optical Theorem thus disallows imaginary parts from logarithms that would appear post integration. This physical constraint is imposed by the factor of $(D-4)$, which absorbs the divergence of $I_2(k_{12})$ and pushes the logarithm to be subleading at $\mathcal{O}(\epsilon)$.

@@

where $g_i \equiv 2 k_{i-1}^\mu F^{\mu\nu}_i k^{\nu}_{i+1}$. While not obvious when expressed in this form, the numerator of the above tensor structure is proportional to the permutation invariant in the denominator, $\mathcal{O}^{F^2F^2}_{(0,2)} -g_1g_2g_3g_4\sim \sigma_3$, and thus it is local by construction. This set of operators completely spans the space of $D$-dimensional permutation invariant photon effective operators, thereby allowing us to write an on-shell effective amplitude as

@@

\item \textbf{Projection}: Project $D$-dimensional tensor structures along defined 4D spinor helicity states. This will elucidate where anomalous matrix elements appear for non-supersymmetric DBIVA, which has a classical conserved $U(1)$ symmetry. 

@@

Before diving into the fully dressed fermion and vector integrals of DBIVA theories, we will begin with an instructive example of NLSM pions through two-loop order. These amplitudes will serve as the scaffolding for constructing $\mathcal{N}=4$ DBIVA integrands using the double copy. As we will demonstrate, an exciting realization of our results is that the loop-level double copy construction of $\mathcal{N}=4$ DBIVA is consistent with the results that we compute via unitarity. This is the first verification of the double-copy construction for multi-loop amplitudes for non-gravitational theories. Furthermore, it suggests that there should exist a color-dual representation of two-loop NLSM integrands, which thus far has evaded identification in the literature. We see finding such a color-dual pion integrand as an exciting direction of future study.   

@@

\mathcal{A}^{\text{NLSM}}_n = \sum_{\sigma \in S^{n-2}} C^{\text{ladder}}_{(1\sigma n)} A^{\text{NLSM}}_{(1\sigma n)}\,.

@@

With this, we are finally prepared to start calculating! As we will see, the $D$-dimensional calculations for pions are quite simple -- the integrand expressions are only a few lines of dot products. Guided by the simplicity of NLSM, we will label each step in the process according to the outline provided at the beginning of the section. While the complexity of the calculation will necessarily increase for DBIVA theories, the general procedure will remain the same.
